Unrecognized States: Theory and Cases

### 2022-02-26 teaching.tex
update and add detail to teaching example

\section{Unrecognized states in context (i.e. the strategic situation)}
Unrecognized states destabilize the international system and impoverish their residents. Because the borders of these territories are contested, the threat of violent conflict is ever present. Unrecognized states are unable to sign trade agreements or receive most foreign aid, and most face economic sanctions, a combination that leaves their residents isolated and impoverished. Thus, unrecognized statehood is a profoundly undesirable outcome, and yet it is nonetheless often a stable equilibrium. The long-standing unrecognized states of Somaliland and the Turkish Republic of Northern Cyprus, as well as many of the unrecognized states that emerged when the Soviet Union collapsed, still enjoy de facto independence a quarter century later. Unrecognized states are thus quintessential intractable conflicts: difficult but not impossible to resolve (e.g. Crocker, Hampson, and Aall 2004; 2005).\\

Because unrecognized statehood is such a clearly unfavorable outcome for the two primary parties involved – the unrecognized state itself and the home state from which it is attempting to secede – much of the existing work has either treated unrecognized states as an aberration or a temporary phenomenon, or, conversely, as utterly intractable conflicts rooted in the irrationality of one or more actors. We use game theoretic analysis to challenge these understandings. We argue that unrecognized statehood can, in fact, be a stable equilibrium outcome, and that it can be sustained even when all the players involved are perfectly informed and behaving rationally. Game theory provides value in this context by laying bare the mechanisms by which unrecognized statehood is sustained as a stable equilibrium outcome, and thus illuminating the ways in which these intractable conflicts can be peacefully transformed. \\

We analyze not only the actions of unrecognized states and the home states from which they are attempting to secede, but also the patrons that support these unrecognized states and the actors in the international community who work to induce peaceful settlement. By formalizing the incentives of these third-party actors and the actions available to them, we are able to analyze the conditions under which their conduct can lead to peaceful resolution, and when it can lead to war or continued stalemate (Buzard, Graham, and Horne 2017). \\

We define unrecognized states as territories in which a non-state actor controls territory, governs a population, and seeks but does not receive broad recognition as an independent state. The newest unrecognized states were formed when the Soviet Union collapsed in the early 1990s. In the case of more recent cases of militarily successful secession, such as South Sudan, unrecognized statehood has been avoided. This suggests that, if the stalemates sustaining the six current unrecognized states (Abkhazia, Nagorno-Karabakh, Somaliland, South Ossetia, Transnistria, and the Turkish Republic of Northern Cyprus) can be successfully resolved, a world without unrecognized states is possible (Buzard, Graham and Horne 2019).\\

@@

\vskip.3in
\subsection{Timing and Information}
The patron makes its investment. This investment is seen by all other players. Then the home government and secessions move simultaneously.

@@

@@

>>>>

\section{Unrecognized states in context (i.e. the strategic situation)}
Unrecognized states destabilize the international system and impoverish their residents. Because the borders of these territories are contested, the threat of violent conflict is ever present. Unrecognized states are unable to sign trade agreements or receive most foreign aid, and most face economic sanctions, a combination that leaves their residents isolated and impoverished. Thus, \textbf{unrecognized statehood is a profoundly undesirable outcome}, and yet it is nonetheless often a stable equilibrium. \textbf{Somaliland and the Turkish Republic of Northern Cyprus are long-standing unrecognized states}, and many of the unrecognized states that emerged when the Soviet Union collapsed still enjoy de facto independence a quarter century later. Unrecognized states are thus quintessential intractable conflicts: difficult but not impossible to resolve (e.g. Crocker, Hampson, and Aall 2004; 2005).\\

Because unrecognized statehood is such a clearly unfavorable outcome for the two primary parties involved – the unrecognized state itself and the home state from which it is attempting to secede – much of the existing work has either treated unrecognized states as an aberration or a temporary phenomenon, or, conversely, as utterly intractable conflicts rooted in the irrationality of one or more actors. We use game theoretic analysis to challenge these understandings. We argue that unrecognized statehood can, in fact, be a stable equilibrium outcome, and that it can be sustained even when all the players involved are perfectly informed and behaving rationally. \textbf{Game theory provides value in this context by laying bare the mechanisms by which unrecognized statehood is sustained as a stable equilibrium outcome, and thus illuminating the ways in which these intractable conflicts can be peacefully transformed.} \\

\textbf{We analyze not only the actions of unrecognized states and the home states from which they are attempting to secede, but also the patrons that support these unrecognized states and the actors in the international community who work to induce peaceful settlement.} By formalizing the incentives of these third-party actors and the actions available to them, \textbf{we are able to analyze the conditions under which their conduct can lead to peaceful resolution, and when it can lead to war or continued stalemate} (Buzard, Graham, and Horne 2017). \\

\textbf{We define unrecognized states as territories in which a non-state actor controls territory, governs a population, and seeks but does not receive broad recognition as an independent state.} In the most recent militarily-successful secession---South Sudan---unrecognized statehood has been avoided. \textbf{This suggests that, if the stalemates sustaining the eight current unrecognized states (Abkhazia, Donetsk, Luhansk, Nagorno-Karabakh, Somaliland, South Ossetia, Transnistria, and the Turkish Republic of Northern Cyprus) can be successfully resolved, a world without unrecognized states is possible (Buzard, Graham and Horne 2019). On the other hand, as the current conflict in Ukraine demonstrates, the ongoing existence of unrecognized states can be profoundly destabilizing (Sommerlad, 2022).} \\

@@

\vskip.3in
\subsection{Timing and Information}
The patron makes its investment. This investment is seen by all other players. Then the home government and secessionists move simultaneously.

@@

\vskip.5in
\subsection{Solution Concept}
This game has a dynamic component, as the government and the secessionists move after seeing the patron's choice. Further, we assume that the players have symmetric information about payoffs; that is, all the players know the payoffs associated with each outcome. We will also assume that the players know each other well enough to be able to correctly predict each other's behavior; this is likely to be the case when the same leaders have been in place for a while so that they have interacted repeatedly. As this is a high-stakes `game,' all the players have strong incentives to seek out information and think carefully about what the other players will do before choosing the strategy that will maximize their expected payoffs, and this is common knowledge amongst the players. Finally, no commitment devices are available to enforce players' promises, so the players cannot make commitments that are incredible. \\

These assumptions (the game is dynamic, information in symmetric, and the actors have common knowledge of sequential rationality and can correctly predict each other's behavior) imply that Subgame Perfect Nash Equilibrium (SPNE) is the appropriate solution concept for this strategic situation (Watson p. 45 and p. 189).

@@

Sommerlad, J. 2022. ``Ukraine crisis: What are the Donetsk and Luhansk People’s Republics?.'' The Independent. February 25, 2022. https://www.independent.co.uk/news/world/europe/ukraine-donetsk-luhansk-region-russia-war-b2023842.html. Accessed February 26, 2022.
